---
title: 'Lab 4: Land-cover pattern and change'
author: "Eric Jensen"
date: "February 13, 2020"
output: html_document
---

```{r setup, include=FALSE}
library(raster)            #for raster data; version 2.6-7 used
library(rasterVis)         #for plotting rasters;  version 0.45 used
library(landscapemetrics)  #for patch, class, and landscape metrics
library(rgdal)             #for raster data, projections; version 1.3-4 used
library(tidyverse)
library(lattice)
library(ggthemes)
```

## Fort Collins Raster
### Question 1
**1. Generate a bar plot showing the proportion of landcover types in the Fort Collins area. Which is most prevalent? Use ggplot for the plotting. See example figure (landcover_of_fort_collins.pdf) (2 pts)** 

```{r}
#The following code could be useful for the Fort Collins raster
nlcd <- raster("C:/Users/erjensen/Documents/ECOL620/Lab4/data_for_lab4/fort_collins.tif")
#nlcd <- as.factor(nlcd)

# Convert raster to dataframe and create columns of percent
nlcd_df_first <- as.data.frame(nlcd)
nlcd_df <- nlcd_df_first %>%
  group_by(fort_collins)%>%
  mutate(count = n()) %>% # Create column with counts of each class
  unique() %>%
  ungroup() %>%
  mutate(sum = sum(count)) %>% #create column with sum of total pixels
  mutate(pctcover = (count/sum)*100) %>% # calculate percent of each cover type
  select(-sum) %>%
  arrange(fort_collins)

  #full_join(land_cover, by ='landcover')#remove sum column

# Add cover type names to the dataframe
nlcd_df[,"landcover"] <- c("Open Water", "Developed, Open Space","Developed, Low Intensity",
                              "Developed, Medium Intensity","Developed, High Intensity",
                              "Barren Land","Deciduous Forest", "Evergreen Forest","Mixed Forest",
                              "Shrub/Scrub","Grassland/Herbaceous","Pasture/Hay","Cultivated Crops",
                              "Woody Wetlands","Emergent Herbaceous Wetlands")

# Add colors to the dataframe
nlcd_df$land_col <- c("#4f6d9f", "#decece", "#d29b85", "#de3021", "#9d1f15", "#b2afa5", "#7aa76d", "#336338",
                      "#c0cb99","#cebb89", "#edecd0","#ddd75c", "#a67538", "#bfd7eb", "#7ba3be")

nlcd_df <- within(nlcd_df, 
                   Position <- factor(fort_collins, 
                                      levels=names(sort(table(fort_collins), 
                                                        decreasing=FALSE))))

# Plot it
ggplot(data = nlcd_df, aes(x=as.factor(fort_collins),y=pctcover, fill = I(land_col)))+
  geom_bar(aes(x = as_factor(landcover)),stat='identity') + # I tried a lot of different ways, but kept having issues with the colors and orders of the classes getting mixed up. This worked
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x='Landcover type', y='Percent class (%)')
remove(nlcd_df,nlcd_df_first)
```

### Question 2
**Calculate the following metrics for Open Water:**
**2. Number of open water patches in the greater Fort Collins region using: 4-neighborhood rule ________ 8-neighborhood rule: _________ (1 pt)**

```{r}
#create a reclassification matrix
nlcd.cat <- unique(nlcd)
nlcd.cat.wat <- c(1,0,0,0,0,0,0,0,0,0,0,0,0,0,0)

reclass.mat <- cbind(nlcd.cat,nlcd.cat.wat)

#water binary layer from reclassification matrix
nlcd.water <- reclassify(nlcd,reclass.mat)

#Number of patches based on rook's and queen's rules
rook_patches <- lsm_c_np(nlcd.water, directions = 4)[2,6]$value
queen_patches <- lsm_c_np(nlcd.water, directions = 8)[2,6]$value
```

**Answer:** Number of open water patches in the greater Fort Collins region using:  
4-neighborhood rule: **`r rook_patches` ** 
8-neighborhood rule: **`r queen_patches`**  


### Question 3
**3. Largest open water patch size in hectare (using the 4-neighborhood rule): _________ (1 pt)**

```{r}
#Now, calculate patch-level metrics:
water_df<-calculate_lsm(nlcd.water, level = "patch", directions = 4,full_name=T) #this yields all the patch metrics
water_area <- water_df %>%
  filter(metric == 'area') %>%
  filter(class == 1)
big_water_area <- max(water_area$value)
```

**Answer:** Largest open water patch size in hectare (using the 4-neighborhood rule): **`r big_water_area`**

### Question 4: 
**4. Mean open water patch size in hectare and km2 (using the 4-neighborhood rule):_________ (1 pt)**

```{r}
avg_water_area <- mean(water_area$value)
```

**Answer:** Mean open water patch size in hectare and km2 (using the 4-neighborhood rule): **`r avg_water_area`**

### Question 5
**5. Range of core area open water patches in hectare (using the 8-neighborhood rule and edge depth of 3): _________ (1 pt)**

```{r}
water_core <- water_df %>%
  filter(metric == 'core') %>%
  filter(class == 1)

range_water_core <- range(water_core$value)
```

**Answer:** Range of core area open water patches in hectare (using the 8-neighborhood rule and edge depth of 3): **`r range_water_core[1]`** - **`r range_water_core[2]`**

### Question 6 
**6. Of the patch-based metrics, which are most dissimilar? Include the top-three most dissimilar. (1 pt)**

```{r}
#Now, calculate patch-level metrics:
patch_metrics=calculate_lsm(nlcd.water, level = "patch", directions = 8,full_name=T) #this yields all the patch metrics

#correlation matrix
show_correlation(data = patch_metrics[,1:6], method = "pearson", label=T)
```

**Answer:** The three most negatively correlated pairs of metrics are:  
* Contiguity and perimeter-to-area ratio (-0.99)  
* Core area index and perimeter-to-area ratio (-0.91)  
* Circumscribing circle and contiguity (-0.42)  

### Question 7
**7. Generate a plot showing the relationship between the number of open water patches (using the 4-neighborhood rule) and grain size. Use 30m, 60m, 120m, 240, and 600m grain resolution. Use the modal rule and generate your plot using ggplot. (4 pts)**
```{r}
# Create dataframe to accumulate values
nlcd_30_grain <- res(nlcd.water)[1]
nlcd_30_patches <- lsm_c_np(nlcd.water, directions = 4)[2,6]$value
patches_df <- nlcd_30_grain %>% as_tibble() %>% rename(Grain=value) %>% mutate(Patches = nlcd_30_patches)

#list of additional grain sizes of interest
grains <- list(60,90,120,150,180,210,240,270,300,400,500,600) #added more grain sizes to better see trends
for(Grain in grains){
  Aggregated <- aggregate(nlcd.water,(Grain/30))
  Patches <- lsm_c_np(Aggregated, directions = 4)[2,6]$value
  bound_df <- as_tibble(cbind(Grain, Patches)) 
  patches_df <- rbind(patches_df, bound_df)
  }

ggplot()+
  geom_point(data = patches_df, aes(x = Grain, y = Patches))+
  labs(x='Grain size (meters)', y='Number of patches') +
  theme_few()
```

#### **Reclassify fort_collins.tif to the scheme outlined on lines 39 to 46. Calculate the following metrics:**

```{r}
#------------------------------------------#
#land-cover type (nlcd original categories)
#1 = forest:41-43
#2 = developed:21-24
#3 = agriculture:81,82
#4 = herbaceous:71-74
#5 = open:31, 51-52
#6 = wetland:90,95
#7 = water:11-12
#------------------------------------------#

#create a reclassification matrix
nlcd.cat <- unique(nlcd)
nlcd.cat.recl <- c(7,2,2,2,2,5,1,1,1,5,4,3,3,6,6)

reclass.mat <- cbind(nlcd.cat,nlcd.cat.recl)

#reclassified nlcd raster
nlcd.recl <- reclassify(nlcd,reclass.mat)
```

### Question 8 
**8. Number of agriculture patches in the greater Fort Collins region using: 4-neighborhood rule ________ 8-neighborhood rule: _________ (1 pt)**

```{r}
ag_rook_patches <- lsm_c_np(nlcd.recl, directions = 4)[3,6]$value
ag_queen_patches <- lsm_c_np(nlcd.recl, directions = 8)[3,6]$value
```

**Answer:** Number of agriculture patches in the greater Fort Collins region using: 4-neighborhood rule **`r ag_rook_patches`** 8-neighborhood rule: **`r ag_queen_patches`**

### Question 9
**9. Mean Euclidean nearest neighbor distance of agriculture patches (Using the 4- neighborhood rule): _________ (1 pt)**

```{r}
ag_rook_enn <- lsm_c_enn_mn(nlcd.recl,directions=4)[3,6]$value
```

**Answer::** Mean Euclidean nearest neighbor distance of agriculture patches (Using the 4- neighborhood rule): **`r ag_rook_enn`**

### Question 10
**10. Include a plot (using the landscape metrics package, e.g., show_lsm) of developed and open water landcover types showing the number of core areas (Using the 4- neighborhood rule). (1 pt)**

```{r}
show_lsm(nlcd.recl, what = "lsm_p_core", direction=4, class = c("2","7"), label_lsm = F, labels = F)
```

### Question 11
**Calculate the following landscape metrics (for the original fort_collins.tif). In addition to quantifying the values, briefly describe the meaning of the values. For instance, what does a high value of contagion indicate?**
**11. Land-cover richness: _________ (1 pts)**

```{r}
richness <- length(unique(values(nlcd)))
```
 
 **Answer:** The landcover richness value is **`r richness`**. This value indicates that there are 15 landcover classes in the Fort Collins NLCD data.

### Question 12
**12. Shannon diversity: _________ (1 pts)**

```{r}
C <- table(values(nlcd))
P <- C / sum(C)
D <- -sum(P * log(P))
E <- D/log(length(C))

lsm_l_shdi(nlcd) # from LSM
print(D) #shannon's diversity index
```

**Answer**: The Shannon diversity index for the Fort Collins NLCD data is **`r D`**. Values of this index are calculated based on the richness of classes as well as the abundance of units (in our case, pixels) in each class. A value of `r D` is intermediate in the spectrum of Shannon values, indicating a somewhat diverse landscape.

### Question 13
**13. Shannon evenness: _________ (1 pts)**

```{r}
print(E) #shannon's evenness index
lsm_l_shei(nlcd) # from LSM
```

**Answer:** The Shannon evenness index for the Fort Collins NLCD data is **`r E`**. Values of this index are calculated as a ratio between the observed Shannon diversity and the theoretical maximum of Shannon diversity. It measures dominance. Shannon values range from 0 (totally dominated by a single class) and 1 (perfectly evenly distributed among the classes), so a value of `r E` indicates a relatively high level of evenness in the dataset.

### Question 14
**14. Contagion: _________ (1 pts)**

```{r}
contag <- lsm_l_contag(nlcd)[1,6]$value
```

**Answer:** The value for contagion is **`r contag`**. Contagion valculates the probability of two randomly selected cells of being in the same class and is affected by both the dispersion (values close to 0) and interspersion (values close to 100) of classes. A value of `r contag` indicates that cells of like classes are fairly clustered in the NLCD, as we would expect with a landcover map with contiguous urban, forest, agricultural, etc. areas.
